#Lab_03 {lab_03}

Natalie Weger, 2/8/2024

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse here
library(tidyverse)

# Load janitor here
library(janitor)

```

### Load data

In this data set, each row represents a city or town in Maryland with the 2010 and 2020 figures for:

-   Total population
-   White only population
-   Black or African American only population
-   Hispanic only population
-   Native American only population
-   Asian only population
-   Pacific Islander (including native Hawiian) only population

Background on this data from [Social Explorer](https://www.socialexplorer.com/data/ACS2016_5yr/metadata/?ds=ACS16_5yr&table=B02001)

```{r}
# Write code to load the Maryland race by city, place for 2010 - 2020 and call it md_race.
# The data is called lab_03.csv, and is in the same folder as lab_03.qmd (this data notebook)

library(readr)
md_race <- read_csv("lab_03.csv")
```

## Answer questions

Q1: Which five cities have the largest overall Hispanic population in 2020? Name the five cities that had the largest Hispanic population in 2020.

A1:

```{r}

# Write your code in this codeblock.

md_race %>% 
  select(place,x2020_hispanic) %>% 
  slice_max(x2020_hispanic, n = 5)

#The top five cities that had the largest Hispanic population in 2020 is Baltimore City, Wheaton CDP, Germantown CDP, Silver Spring CDP and Aspen Hill CDP. 


```

Q2: Find the cities with the largest Black population and arrange the results in descending order.

A2:

```{r}
#write your code here

md_race %>% 
  select(place, x2020_black) %>% 
  arrange(desc(x2020_black))

#The city with the largest Black population in 2020 is Baltimore City with a population of 375002. The second largest is Waldorf CDP with a population of 47903 and the third largest is Bowie City with a population of 32784. 

```

Q3: Create a table called top_hispanic that shows the top 15 communities with the largest Hispanic population in 2020, and also display the 2010 Hispanic population and the total community population for 2020. Look at your results - what stands out to you? Craft a one-paragraph story pitch for your editor.

A3:

```{r}

top_hispanic <- md_race %>% 
  select(place,x2020_hispanic,x2010_hispanic,x2020_total) %>% 
  slice_max(x2020_hispanic, n=15)
View(top_hispanic)

#What stood out to me is that the total Hispanic population between 2010 and 2020 had increased by about at least 1,000 in all 15 cities with the largest Hispanic population in 2020. I also find it interesting that the city with the largest Hispanic population in 2020 (Baltimore City) has less than 5% of Hispanic people making up its population, while the second and third city of Wheaton and Germantown has about 44% and 23% of an Hispanic population respectively. The questions that I have about this include: Why and how did the Hispanic population increase for all 15 cities over the past decade? What about Wheaton and Germantown makes it have a higher Hispanic population percentage? To start answering these questions, I would start with Dana Moore, the chief equity officer in Baltimore City and Mark Elrich, the county executive in Montgomery County. I would also interview civilians within these cities, specifically within the first three, who have been living in the cities for more than a decade to see if they noticed anything interesting of the population changing. 
```
