---
title: "w"
format: html
editor: visual
name: natalie weger
date: 3/12/2024 
project: data journalism notebook final
---

Natalie Weger

3/12/2024

Data Journalism Notebook Final

## Question 1:

Are there extreme differences between different communities in Baltimore (in 2021) in the percentage of white students enrolled in school?

```{r}

#loading librarires
library(tidyverse)
library(rio)
library(janitor)

#rsw
#rsw - i can't load the data with file paths like these - it's directing me to your hard drive. the data wasn't on your github either

#downloading the data 
baltimore_white <- read_csv("white_students.csv")

#summarizing the data to find differences in baltimore neighborhoods 
summary(baltimore_white$wstud21)

#finding the top 10 baltimore communites with highest white student population 
baltimore_white %>% 
  select(CSA2020, wstud21) %>% 
  slice_max(wstud21, n = 10) %>% 
  arrange(desc(wstud21))
  
#finding the 10 baltimore communities with the lowest white student population 
baltimore_white %>% 
select(CSA2020, wstud21) %>% 
  slice_min(wstud21, n = 10) %>% 
  arrange(wstud21)

```

Answer #1: There does seem to be extreme differences between communities in the amount of white students who lived in Baltimore in 2021. For example, South Baltimore was made up of about 80.51% of white students and Canton was made up of about 67.86% of white students. On the other hand, there are some communities, such as Edmondson Village and Greater Rosemont, that have less than 1% white students. This makes me wonder what are the differences between these neighborhoods, despite all of them being in Baltimore, that could cause this difference.

It is also interesting how Baltimore seemed to have a low percentage of white students in 2021. For example, the median of the data was about 4.42%, meaning that half of the communities in Baltimore have less than 4.42% of white students. Additionally, the third quartile of white students was about 13.18%, while the maximum neighborhood of South Baltimore was at 80.51%. It's important to look into why most of Baltimore seems to have a low percentage of white students, which inversely means that most school communities have a high percentage of people of color.

## Question 2:

Does this difference of white students cause any differences in the quality of education, such as reading test scores? Since there is only data available from 2014, was there any comparisons from then?

```{r}

#rsw - i can't load the data with file paths like these - "~/Downloads/ - it's directing me to your hard drive. the data wasn't on your github either



#download reading test scores 
reading_scores <- read.csv("test_scores.csv")

#finding the top 5 baltimore communites with highest reading scores
reading_scores %>% 
  select(Community.Statistical.Area..CSA., X2014) %>% 
  slice_max(X2014, n = 5) %>% 
  arrange(desc(X2014))

#finding the top 5 baltimore communites with lowest reading scores
reading_scores %>% 
  select(Community.Statistical.Area..CSA., X2014) %>% 
  slice_min(X2014, n = 5) %>% 
  arrange(X2014)

#finding the top 5 baltimore communites with highest white student population in 2014 
baltimore_white %>% 
  select(CSA2020, wstud14) %>% 
  slice_max(wstud14, n = 5) %>% 
  arrange(desc(wstud14))

#finding the 5 baltimore communities with the lowest white student population in 2014
baltimore_white %>% 
select(CSA2020, wstud14) %>% 
  slice_min(wstud14, n = 5) %>% 
  arrange(wstud14)

#creating top 5 communities with highest reading score table 
top_5_communities <- reading_scores %>%
  select(Community.Statistical.Area..CSA., X2014) %>%
  slice_max(X2014, n = 5) %>%
  arrange(desc(X2014))
  
#creating a chart of my above table with the top 5 baltimore communites with the highest reading scores 
top_5_communities %>%
  ggplot() +
  coord_flip() + 
  geom_col(aes(x = reorder(Community.Statistical.Area..CSA., X2014), y = X2014)) +
  labs(
    title = "Percent of 8th grade students passing MSA reading in 2014",
    x = "Communities",
    y = "Percent of Students",
    caption = "Source: Baltimore Neighborhood Indicator Alliance\nBy Natalie Weger\n03/12/2024"
  )


```

Answer #2: Looking at data of 55 communities, it does seem that the race in Baltimore communities did play a factor into how students in Baltimore communities did on MSA reading exams from 2014. The communities that have performed the highest in MSA reading include Greater Roland Park/Poplar Hill with about a 94.29% passing rate, North Baltimore/Guilford/Homeland with about a 90.48% passing rate, South Baltimore with about a 85% passing rate and then also Cross-Country/Cheswolde and Mount Washington/Cold Spring. Of the top five Baltimore communities that have performed well on MSA reading, three of them, North Baltimore/Guilford/Homeland, Greater Roland Park and South Baltimore were also among the top five counties with the highest white student population in 2014. It was notable that South Baltimore had the highest white student population and the third highest MSA passing rate of 85%.

I also found the Baltimore communities with the lowest passing rate of MSA reading in 2014, including Cherry Hill with about 42.15%, Upton/Druid Heights with about 42.15% and The Waverlies with about 42.25%, Westpot/Mount Winans/Lakeland and Sandtown-Winchester/Harlem Park. Of these five communities, two also saw the lowest white percentage of students, which was Upton/Druid Heights of about 0.39% of white students and Sandtown-Winchester/Harlem park with 0.37% of white students. This seems to be a significant finding considering there are 55 Baltimore communities. This data shows that there does seem to be a correlation between the students being white and how the performed in MSA reading in 2014, though there could be other factors that we are not considering.

## Question 3:

Have any communities in Baltimore seen a significant increase or decrease in white students from 2010 to 2020?

```{r}

#create a column of the percentage change of the white population from 2010-2020
baltimore_white_percent <- baltimore_white %>%
  select(CSA2020, wstud11, wstud21) %>%
    mutate(diff_pct_wstud = (wstud21-wstud11)/wstud11) %>% 
    arrange(desc(diff_pct_wstud))
view(baltimore_white)

#find the top 25 communities with the biggest white population increase 
baltimore_white_percent %>%
  select (CSA2020, diff_pct_wstud) %>%
  slice_max (diff_pct_wstud, n = 25) 
view(baltimore_white)

#find the 25 communities with the biggest white population decrease 
baltimore_white_percent %>%
  select (CSA2020, diff_pct_wstud) %>%
  slice_min (diff_pct_wstud, n = 25) 
view(baltimore_white)
```

Answer #3: Communities in Baltimore do not seem to have a significant increase or decrease in white students from 2011 to 2021. The places with the most significant percentage change of white students includes Midway/Coldstream with a 6.35% increase and Southern Park Heights by 4.20%. Inversely, this means that the students of color in those communities has decreased by 6.35% and 4.20%. There do not seem to be significant decreases in white students in Baltimore communities, with the biggest decrease over the decade being about -0.71% in Greektown/Bayview. This data is important to note because it means that the diversity rates in terms of white students and students of color has not shifted substantially throughout the past 10 years. It's important to understand why these rate have not significantly changed, especially considering that the percentage of white students in Baltimore seems to be extremely low.

## Reporting Memo: 

Based on my data analysis for this project, my biggest finding is that there does seem to be a correlation between the amount of white students in Baltimore and the number of 8th grade students who passed MSA reading in 2014. Although correlation doesn't equal causation, I think that the relationship is something worth looking into. Of the top five Baltimore communities that had the highest passing rate of MSA reading for 8th grade students, three of them (North Baltimore/Guilford/Homeland, Greater Roland Park and South Baltimore) also had the highest white student population. Since there are 55 communities in Baltimore it is intriguing that three of the communities would show up in these results. I would contact the North Baltimore Local Schools Board of Education, Guilford Public Schools department of communications, Greater Roland Park education communications department and the Baltimore City Public Schools education department to see if they have any comments about why the three of the five Baltimore communities that have the highest white student population also had the highest MSA passing rate in 2014. Other interesting findings of this piece is that South Baltimore does seem to be an outlier in the amount of white students within its public schools. This is because South Baltimore was made up of about 80.21%  of white students in 2021, while the median of the data was about 4.42%, meaning that half of the communities in Baltimore have less than 4.42% white students. This is interesting, considering that location-wise, South Baltimore is still towards the middle of Baltimore. For this data, I would contact Colline Emmanuelle, who is a communications manager in South Baltimore. Lastly, another piece of data that I found interesting is that the white student population (and therefore the minority population) in Baltimore has changed over the years. For this information, I would contact the mayor of Baltimore, Brandon Scott, to comment on this trend and the insights found within this data.

**Tweet #1:** Of the five Baltimore communities with the highest white student population, three of those communities had the highest MSA reading passing rate in 2014. 

North Baltimore/Guilford/Homeland, Greater Roland Park and South Baltimore were in the top five of 55 Baltimore communities that had the highest MSA reading passing rate. 

For \@newspaper:

\*link\*

**Tweet #2:** South Baltimore had the highest white student population in 2021 with about 80.51% white students in 2021. However, half of the communities in Baltimore were made up of less than 4.42% of white students in that same year.

Learn more about why South Baltimore has a higher white student population in my latest for \@newspaper:

\*link\*
